---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
library(curatedMetagenomicData)
```

# curatedMetagenomicData

<!-- badges: start -->
[![code quality](https://img.shields.io/codefactor/grade/github/waldronlab/curatedMetagenomicData)](https://www.codefactor.io/repository/github/waldronlab/curatedmetagenomicdata)
[![coverage](https://img.shields.io/codecov/c/github/waldronlab/curatedMetagenomicData)](https://codecov.io/gh/waldronlab/curatedMetagenomicData)
<!-- badges: end -->

The `r BiocStyle::Biocpkg("curatedMetagenomicData")` package provides standardized, curated human microbiome data for novel analyses. It includes marker abundance, marker presence, and relative abundance for samples collected from different body sites, with gene families, pathway abundance, and pathway coverage reinstated as they are added to the cMD4 catalog. The bacterial, fungal, and archaeal taxonomic abundances for each sample were calculated with [MetaPhlAn3](https://github.com/biobakery/MetaPhlAn), and metabolic functional potential was calculated with [HUMAnN3](https://github.com/biobakery/humann). The manually curated sample metadata and standardized metagenomic data are available as (Tree)SummarizedExperiment objects.

## What's new in cMD 4.0 ("cMD4")

Starting with version 4.0.0, `r BiocStyle::Biocpkg("curatedMetagenomicData")` serves sample-level profiling data from the published cMD4 DuckDB catalog of hive-partitioned parquet files produced by `curatedMetagenomicDataETL`, replacing the cMD3 ExperimentHub backend. The user-facing API is unchanged: `curatedMetagenomicData()`, `returnSamples()`, and `mergeData()` keep their signatures (including `dryrun`, `counts`, and `rownames`) and still return `(Tree)SummarizedExperiment` objects, so existing user code continues to work.

Key changes:

- **Harmonized metadata objects.** The legacy `sampleMetadata` object has been replaced by two `data.frame` exports shipped directly with the package: `harmonized_meta` (one row per sample, restricted to columns defined in the cMD harmonization schema) and `all_meta` (the same rows plus study-specific columns still under harmonization). Both expose `study_name` and `sample_id` for use with `curatedMetagenomicData()` and `returnSamples()`.
- **Catalog URL.** Defaults to `https://minio.cancerdatasci.org/cmgd-export/cmgd.duckdb`; override with `options(curatedMetagenomicData.duckdb_url = ...)`.
- **Resource titles.** Still use the cMD3 dotted format `<runDate>.<study_name>.<dataType>`. Because the cMD4 ETL produces a single dated snapshot, `<runDate>` is a fixed release tag (`"cmd4"` by default, override with `options(curatedMetagenomicData.run_date = ...)`).
- **HUMAnN3 data types temporarily unavailable.** `gene_families`, `pathway_abundance`, and `pathway_coverage` are not yet in the cMD4 catalog and currently raise an error if requested; they will be reinstated as soon as they are added to the ETL output.
- **Runtime dependencies.** The runtime package no longer depends on `curatedMetagenomicDataCuration`, `ExperimentHub`, or `AnnotationHub`.

See `NEWS.md` and the vignette for the full set of changes.

## Installation

To install `r BiocStyle::Biocpkg("curatedMetagenomicData")` from Bioconductor, use `r BiocStyle::CRANpkg("BiocManager")` as follows.

```{r, eval = FALSE}
BiocManager::install("curatedMetagenomicData")
```

To install `r BiocStyle::Biocpkg("curatedMetagenomicData")` from GitHub, use `r BiocStyle::CRANpkg("BiocManager")` as follows.

```{r, eval = FALSE}
BiocManager::install("waldronlab/curatedMetagenomicData", dependencies = TRUE, build_vignettes = TRUE)
```

Most users should simply install `r BiocStyle::Biocpkg("curatedMetagenomicData")` from Bioconductor.

## Sample metadata

`r BiocStyle::Biocpkg("curatedMetagenomicData")` ships two curated sample metadata tables that users typically explore before pulling assay data:

```{r, eval = FALSE}
# schema-only, portable cohort assembly
harmonized_meta

# schema plus study-specific columns still under harmonization
all_meta
```

Both tables include `study_name` (used with `curatedMetagenomicData()` to query resources) and `sample_id` (used with `returnSamples()` to pull samples across studies).

## Examples

To access curated metagenomic data, users will use the `curatedMetagenomicData()` function both to query and return resources. Multiple resources can be queried or returned with a single call, but only the titles of resources are returned by default.

```{r, collapse = TRUE}
curatedMetagenomicData("AsnicarF_20.+")
```

When the `dryrun` argument is set to `FALSE`, a `list` of `SummarizedExperiment` and/or `TreeSummarizedExperiment` objects is returned. The `rownames` argument determines the type of `rownames` to use for `relative_abundance` resources: either `"long"` (the default), `"short"` (species name), or `"NCBI"` (NCBI Taxonomy ID). When a single resource is requested, a single-element `list` is returned.

```{r, collapse = TRUE, message = FALSE}
curatedMetagenomicData("AsnicarF_2017.relative_abundance", dryrun = FALSE, rownames = "short")
```

When the `counts` argument is set to `TRUE`, relative abundance proportions are multiplied by read depth and rounded to the nearest integer prior to being returned. When multiple resources are requested, the `list` contains named elements corresponding to each `SummarizedExperiment` and/or `TreeSummarizedExperiment` object.

```{r, collapse = TRUE, message = FALSE}
curatedMetagenomicData("AsnicarF_20.+.relative_abundance", dryrun = FALSE, counts = TRUE, rownames = "short")
```

For cross-study cohort assembly, filter `harmonized_meta` (or `all_meta`) to the samples of interest and pass the result to `returnSamples()`:

```{r, eval = FALSE}
library(dplyr)

harmonized_meta |>
    filter(age_years >= 18, !is.na(bmi), body_site == "feces") |>
    returnSamples("relative_abundance", rownames = "short")
```

## Analyses

See [curatedMetagenomicAnalyses](https://github.com/waldronlab/curatedMetagenomicAnalyses) for analyses in R and Python using `r BiocStyle::Biocpkg("curatedMetagenomicData")`.

## Contributing

To contribute to the `r BiocStyle::Biocpkg("curatedMetagenomicData")` R/Bioconductor package, first read the [contributing guidelines](CONTRIBUTING.md) and then open an issue. Also, note that in contributing you agree to abide by the [code of conduct](CODE_OF_CONDUCT.md).

---

Pasolli E, Schiffer L, Manghi P, Renson A, Obenchain V, Truong D, Beghini F, Malik F, Ramos M, Dowd J, Huttenhower C, Morgan M, Segata N, Waldron L (2017). Accessible, curated metagenomic data through ExperimentHub. *Nat. Methods*, **14** (11), 1023-1024. ISSN 1548-7091, 1548-7105, doi: [10.1038/nmeth.4468](https://doi.org/10.1038/nmeth.4468).
