---
title: "Paper Figures"
abstract: >
    The following script reproduces all results shown in our manuscript
    "Accessible, curated metagenomic data through ExperimentHub". Please note 
    that much of this analysis occurred during the development stages of 
    curatedMetagenomicData and was written by multiple authors. As such, while 
    this analysis is methodologically sound, there may now exist more efficient 
    ways to produce the same results syntactically. For example, the `curatedMetagenomicData()`
    convenience function provides a simpler way to access the datasets, also providing 
    access to multiple versions of the same dataset if they exist. See `?curatedMetagenomicData`
    for more information.
output:
    BiocStyle::html_document
vignette: >
    %\VignetteIndexEntry{Paper Figures}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.show = "hold")
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

# Setup

Before running the script, please note that:

* Reproducing the results of our manuscript requires Bioconductor >= 3.6, which will be the development version until October 2017. See http://bioconductor.org/developers/how-to/useDevel/ for instructions on installing the development branch of Bioconductor.

* The classification analyses (Figure 1 Example 1 and Supplemental Figure 1) create very large matrices that can exceed default limits for R pointer size and the C stack limit on some systems. If you see `Error: protect(): protection stack overflow` or `Error: segfault from C stack overflow`, one of these is the issue. From GNU/Linux, these limits can be increased from the command-line as follows:

   ```
   ulimit -s 65535   #addresses C stack overflow
   R --vanilla --max-ppsize=500000    #addresses protect() error
   ```
   
   If you are running on a machine with limited memory, a 32-bit installation of R, otherwise have trouble with stack overflows, or you do not want to run these classification examples, change `eval=TRUE` to `eval=FALSE` in the chunk options for the chunks named `fig1eg` and `suppfig1` to skip these analyses.

* For the sake of (quick) reproducibility, the classification analyses (Figure 1 Example 1 and Supplemental Figure 1) are conducted here by considering fixed values of "ntree" and "mtry". Better accuracies may be obtained by tuning these parameters (such as through grid search), although we observed similar results associated with a much higher computational load.

* Several hours are needed to reproduce Supplemental Figure 3.

* When running Supplemental Figure 5 a warning occurs that "The data you have provided does not have any singletons." This is because the estimated relative abundance output of MetaPhlAn2 (and other metagenomic profiling software), even when converted to approximate counts by multiplying by read depth, does not include singletons. It is not a result of trimming, and does not affect the validity of the calculated Shannon diversity measure in as much as any relative abundance data, such as microbiome data from metagenomics, can be used to estimate alpha diversity. However, we recommend that the chao1 diversity measure not be used (Haegeman et al, The ISME Journal (2013) 7, 1092–1101; doi:10.1038/ismej.2013.10).

Then enter the following from within R to build the html:

```{r, eval=FALSE}
library(rmarkdown)
rmarkdown::render("PaperFigures.Rmd", "html_document")
```

# Make sure a compatible version of Bioconductor is in use

```{r}
if(!suppressPackageStartupMessages(require(Biobase))){
  stop("Please install Bioconductor, see www.bioconductor.org/install")
}
(ExperimentHubIsCurrent <- Biobase::package.version("ExperimentHub") >= "1.3.0")
if(!ExperimentHubIsCurrent){
  stop("This analysis currently requires the development branch of Bioconductor. See http://bioconductor.org/developers/how-to/useDevel/")
}
```

# Run

```{r libraries, include=TRUE, cache=FALSE}
suppressPackageStartupMessages({
    library(curatedMetagenomicData)
    library(phyloseq)
    library(randomForest)
    library(caret)
    library(pROC)
    library(gclus)
    library(doMC) #not supported in Windows
    library(stats)
    library(cluster)
    library(fpc)
    library(reshape2)
})
```

Set the number of cores to use:
```{r, cache=FALSE}
registerDoMC(32) #not supported in Windows
```

# Coordinated Color Scheme

```{r colors}
blue <- "#3366aa"
blueGreen <- "#11aa99"
green <- "#66aa55"
paleYellow <- "#cccc55"
gray <- "#e5e5e5"
purple <- "#992288"
red <- "#ee3333"
orange <- "#ee7722"
yellow <- "#ffee33"
darkGray <- "#b3b3b3"
brown <- "#655643"
lightBlack <- "#777777"
black <- "#000000"
pallet <- c(blue, blueGreen, green, paleYellow, gray, purple, red,
            orange, yellow, darkGray, brown, lightBlack, black)
pallet_reduced <- c(blue, green, darkGray, purple, red, black)
n <- length(pallet)
image(1:n, 1, as.matrix(1:n), col = pallet, xlab = "", ylab = "", xaxt = "n", 
      yaxt = "n", bty = "n")
```

# Figure 1, Example 1: Classification

```{r fig1eg1, fig.height=8, fig.width=8, eval=TRUE}
dataset_list <-
    c(  "KarlssonFH_2013 (T2D)",
        "LeChatelierE_2013 (Obesity)",
        "NielsenHB_2014 (IBD)",
        "QinJ_2012 (T2D)",
        "QinN_2014 (Cirrhosis)",
        "ZellerG_2014 (CRC)"
     )
class_list <- c("T2D", "obesity", "IBD", "T2D", "cirrhosis", "CRC")
data_list <- matrix(nrow = 5, ncol = length(dataset_list))
## update
data_list[1, ] <- c("EH439", "EH547", "EH457", "EH475", "EH541", "EH535")  # Species abundance
data_list[2, ] <- c("EH440", "EH548", "EH458", "EH476", "EH542", "EH536")  # Pathway abundance
data_list[3, ] <- c("EH441", "EH549", "EH459", "EH477", "EH543", "EH537")  # Pathway coverage
data_list[4, ] <- c("EH437", "EH545", "EH455", "EH473", "EH539", "EH533")  # Marker abundance
data_list[5, ] <- c("EH438", "EH546", "EH456", "EH474", "EH540", "EH534")  # Marker presencee

## update
eh <- ExperimentHub()

set.seed(0);

for (j in 1:length(data_list[, 1])) {
  for (i in 1:length(dataset_list)) {
    filename <- paste0(paste("rf", j, i, sep = "_"), ".rds")
    message(paste("Training:", j, i, dataset_list[i]))
    if(file.exists(filename)) next
    taxabund <- eh[[data_list[j, i]]]
    
    feat <- t(exprs(taxabund))
    if (j == 1) {
      feat <- feat[, grepl("(s__|unclassified)", colnames(feat))]
      feat <- feat[, !grepl("t__", colnames(feat))]
    }
    if (j <= 3) {
      feat <- predict(preProcess(feat, method=c("zv", "scale", "center")), feat) 
    } else {
      feat <- predict(preProcess(feat, method=c("nzv", "scale", "center"), freqCut = 75/25, uniqueCut = 100), feat)
    }

    meta <- pData(taxabund)
    if (i == 2) {
        meta$study_condition[meta$BMI <= 25] = "lean"
        meta$study_condition[meta$BMI >= 30] = "obesity"
    }
    if (i == 6) {
        meta$study_condition[meta$disease_subtype == "smalladenoma"] = "control"
    }
    meta <- meta["study_condition"]
    all <- cbind(feat, meta)
    if (i == 1) {
        all <- subset(all, study_condition != "IGT")
    }
    if (i == 2) {
        all <- subset(all, study_condition != "control")
    }
    if (i == 4) {
        all <- subset(all, !is.na(study_condition))
    }
    if (i == 6) {
        all <- subset(all, study_condition != "adenoma")
    }

    object <- 
      train(
        study_condition ~ .,
        data = all,
        method = "rf",
        ntree = 500,
        #tuneGrid = expand.grid(.mtry = c(10, 20, 30, 40, 50, 60, 70, 80, 90, 100, 125, 150, 175, 200, 300, 400, 500)),
        tuneGrid = expand.grid(.mtry = c(50)),
        trControl = trainControl(
          method = "repeatedcv",
          number = 10,
          search = "grid",
          summaryFunction = twoClassSummary,
          classProbs = TRUE,
          savePredictions = TRUE
        )
      )
    saveRDS(object, file=filename)
  }
}

message("Finished training, starting plotting.")

ci_bugs <- c()
ci_pab <- c()
ci_pcov <- c()
ci_mab <- c()
ci_mpr <- c()

for (i in 1:length(dataset_list)) {
  message(paste(i, ":", dataset_list[i]))
  rf <- readRDS(paste0(paste("rf", 1, i, sep = "_"), ".rds"))
  ci_bugs <- c(ci_bugs, auc(rf$pred$obs[rf$pred$mtry == rf$bestTune$mtry], rf$pred[, class_list[i]][rf$pred$mtry == rf$bestTune$mtry]))
  
  rf <- readRDS(paste0(paste("rf", 2, i, sep = "_"), ".rds"))
  ci_pab <- c(ci_pab, auc(rf$pred$obs[rf$pred$mtry == rf$bestTune$mtry], rf$pred[, class_list[i]][rf$pred$mtry == rf$bestTune$mtry]))
  
  rf <- readRDS(paste0(paste("rf", 3, i, sep = "_"), ".rds"))
  ci_pcov <- c(ci_pcov, auc(rf$pred$obs[rf$pred$mtry == rf$bestTune$mtry], rf$pred[, class_list[i]][rf$pred$mtry == rf$bestTune$mtry]))
  
  rf <- readRDS(paste0(paste("rf", 4, i, sep = "_"), ".rds"))
  ci_mab <- c(ci_mab, auc(rf$pred$obs[rf$pred$mtry == rf$bestTune$mtry], rf$pred[, class_list[i]][rf$pred$mtry == rf$bestTune$mtry]))
  
  rf <- readRDS(paste0(paste("rf", 5, i, sep = "_"), ".rds"))
  ci_mpr <- c(ci_mpr, auc(rf$pred$obs[rf$pred$mtry == rf$bestTune$mtry], rf$pred[, class_list[i]][rf$pred$mtry == rf$bestTune$mtry]))
}

ci <- data.frame(ci_bugs, ci_pab, ci_pcov, ci_mab, ci_mpr)
colnames(ci) <-
  c(
    "Species abundance",
    "Pathway abundance",
    "Pathway coverage",
    "Marker abundance",
    "Marker presence"
  )
ci.r <- abs(cor(ci))
ci.col <- dmat.color(ci.r)
ci.o <- order.single(ci.r)

par(oma = c(4, 1, 1, 1))

cpairs(
  ci,
  ci.o,
  panel.colors = ci.col,
  col = alpha(pallet_reduced, 0.75),
  pch = 16,
  cex = 2,
  gap = 0.5,
  main = "AUC"
)

par(
  fig = c(0, 1, 0, 1),
  oma = c(0, 0, 0, 0),
  mar = c(0, 0, 0, 0),
  new = TRUE
)
plot(0, 0, type = "n", bty = "n", xaxt = "n", yaxt = "n")
legend("bottom", xpd = TRUE, inset = c(0, 0), bty = "n", pch = 16, cex = 0.6, ncol=3, dataset_list, col = pallet_reduced)
```

# Figure 1, Example 2: Clustering

```{r fig1eg2, fig.height=8, fig.width=8}
joinWithRnames <- function(obj, FUN = I) {
    mylist <- lapply(obj, function(x) {
        df <- data.frame(FUN(x))
        df$rnames28591436107 <- rownames(df)
        return(df)
    })
    bigdf <- Reduce(full_join, mylist)
    rownames(bigdf) <- make.names(bigdf$rnames28591436107)
    bigdf <- bigdf[, !grepl("^rnames28591436107$", colnames(bigdf))]
    return(bigdf)
}
## update
source("https://raw.githubusercontent.com/waldronlab/presentations/master/Waldron_2016-06-07_EPIC/metaphlanToPhyloseq.R")

## update
eh <- ExperimentHub()
myquery <- query(eh, "curatedMetagenomicData")
version <- "20170526."

myquery.stool <- myquery[grepl("stool", myquery$title) & grepl("bugs", myquery$title) & grepl(version, myquery$title) & !grepl("VatanenT_2016", myquery$title) & !grepl("AsnicarF_2017", myquery$title), ]

eset.list <- lapply(names(myquery.stool), function(x) myquery.stool[[x]])

names(eset.list) <- gsub("-", "_", gsub(version, "", myquery.stool$title))

for (i in 1:length(eset.list)) {
    colnames(eset.list[[i]]) <- paste(names(eset.list)[[i]], colnames(eset.list[[i]]), sep = ".")
    pData(eset.list[[i]]) <- pData(eset.list[[i]])[, !sapply(pData(eset.list[[i]]), function(x) all(is.na(x)))]
    eset.list[[i]]$subjectID <- as.character(eset.list[[i]]$subjectID)
}

for (i in seq_along(eset.list)) {
    eset.list[[i]] <- eset.list[[i]][grep("t__", rownames(eset.list[[i]]), invert = TRUE), ]
    eset.list[[i]] <- eset.list[[i]][grep("s__|_unclassified", rownames(eset.list[[i]]), perl = TRUE), ]
}

pdat <- joinWithRnames(eset.list, FUN = pData)
pdat$study <- sub("\\..+", "", rownames(pdat))
ab <- joinWithRnames(eset.list, FUN = exprs)
ab[is.na(ab)] <- 0
eset <- ExpressionSet(assayData = as.matrix(ab), phenoData = AnnotatedDataFrame(pdat))

## update
pseq <- metaphlanToPhyloseq(tax = exprs(eset), metadat = pData(eset), split = ".")

dist_bray <- distance(pseq, method = "bray")

ord_bray <- ordinate(pseq, method = "PCoA", distance = dist_bray)

sample_data(pseq)$bray_cluster_2 <- factor(pam(dist_bray, k = 2, cluster.only = TRUE))


Prev <- as.numeric(otu_table(pseq)["s__Prevotella_copri", ])
sample_data(pseq)$Prevotella_copri <- Prev


pc1 <- ord_bray$vectors[, 1]
pc2 <- ord_bray$vectors[, 2]

otu_tax <- attr(otu_table(pseq), "dimnames")[[1]]
otu_bacteroides <- otu_table(pseq)[grep("s__Bacteroides", otu_tax), ]
sum_bacteroides <- apply(otu_bacteroides, 2, sum)

df_ordinate <- data.frame(pc1, pc2, bact = sum_bacteroides, prev = Prev, bray2 = sample_data(pseq)$bray_cluster_2)
df_bact <- df_ordinate[df_ordinate$bray2 == 2, ]
df_prev <- df_ordinate[df_ordinate$bray2 == 1, ]



ggplot() +
  theme(panel.background = element_rect(fill = "white",
                                        colour = "white",
                                        linetype = "solid"),
        panel.grid.major = element_line(size = 0.5, linetype = 'solid',
                                        colour = gray)) +
  geom_point(data = df_prev, aes(x = pc1, y = pc2, shape = factor(bray2), fill = prev, color=bact), size = 2, alpha=0.5, stroke=1) +
  scale_fill_gradient(low = "white", high = purple) +
  geom_point(data = df_bact, aes(x = pc1, y = pc2, shape = factor(bray2), color = bact), size = 3, alpha=0.5) +
  scale_color_gradient(low = gray, high = "black") +
  labs(x = "Axis 1", y = "Axis 2", title = "PCoA on species abundance, displaying 2 clusters") +
  theme(axis.ticks = element_blank(), axis.text = element_blank(), 
        legend.box = "vertical", legend.position = c(0.175, 0.125), legend.key = element_rect(fill="white"), 
        plot.title = element_text(hjust = 0.5)) +
  scale_shape_manual(values=c(21,18), name="Cluster") +
  guides(shape = guide_legend(ncol=1, labels=c(1,2), order=1),
         fill = guide_colorbar(direction = "horizontal", title = "Prevotella copri"),
         colour = guide_colorbar(direction = "horizontal", title = "Bacteroides"))

```

# Figure 1, Example 3: Abundance across samples

```{r fig1eg3, fig.height=8, fig.width=12}
## update
eh <- ExperimentHub()
myquery <- query(eh, "curatedMetagenomicData")
version <- "20170526."

myquery.stool <- myquery[grepl("stool", myquery$title) & grepl("bugs", myquery$title) & grepl(version, myquery$title) & !grepl("VatanenT_2016", myquery$title) & !grepl("AsnicarF_2017", myquery$title), ]

eset.list <- lapply(names(myquery.stool), function(x) myquery.stool[[x]])

names(eset.list) <- gsub("-", "_", gsub(version, "", myquery.stool$title))

for (i in 1:length(eset.list)) {
    colnames(eset.list[[i]]) <- paste(names(eset.list)[[i]], colnames(eset.list[[i]]), sep = ".")
    pData(eset.list[[i]]) <- pData(eset.list[[i]])[,!sapply(pData(eset.list[[i]]), function(x) all(is.na(x)))]
    eset.list[[i]]$subjectID <- as.character(eset.list[[i]]$subjectID)
}

for (i in seq_along(eset.list)) {
    eset.list[[i]] <- eset.list[[i]][grep("t__", rownames(eset.list[[i]]), invert = TRUE),]
    eset.list[[i]] <- eset.list[[i]][grep("s__|_unclassified", rownames(eset.list[[i]]), perl = TRUE),]
}

pdat <- joinWithRnames(eset.list, FUN = pData)
pdat$study <- sub("\\..+", "", rownames(pdat))
ab <- joinWithRnames(eset.list, FUN = exprs)
ab[is.na(ab)] <- 0
eset_bugs <- ExpressionSet(assayData = as.matrix(ab), phenoData = AnnotatedDataFrame(pdat))

## update
pseq <- metaphlanToPhyloseq(tax = exprs(eset_bugs), metadat = pData(eset_bugs), split = ".")

glom <- tax_glom(pseq, taxrank = "Phylum")

top8phyla <- names(sort(taxa_sums(glom), TRUE)[1:8])
phyla8_subset <- prune_taxa(top8phyla, glom)

phyla_to_sort <-
    data.frame(
        id = 1:8,
        phyla = as.character(tax_table(phyla8_subset)[, "Phylum"]),
        otu = as.character(taxa_names(phyla8_subset))
    )
rownames(phyla_to_sort) <- phyla_to_sort$otu

phylum_ranks <-
    phyla8_subset %>%
    otu_table %>%
    rowSums %>%
    sort(TRUE) %>%
    names

phyla_to_sort <- phyla_to_sort[phylum_ranks, ]

prop <- transform_sample_counts(phyla8_subset, function(i) i/sum(i))

bardat <-
    psmelt(prop) %>%
    select(OTU, Sample, Abundance) %>%
    mutate(Sample = as.numeric(factor(Sample)), OTU = factor(OTU, levels = phyla_to_sort$otu, labels = phyla_to_sort$phyla))

firmicutes_order <-
    bardat %>%
    filter(OTU == "Firmicutes") %>%
    arrange(Abundance) %>%
    select(Sample)

bardat %<>%
    mutate(Sample = as.numeric(factor(Sample, levels = factor(firmicutes_order$Sample)))) %>%
    arrange(desc(OTU), Abundance)

set.seed(14)

bardat %>%
    ggplot(aes(x = Sample, y = Abundance, fill = OTU)) +
    geom_area() +
    scale_fill_manual(values = sample(c(blue, yellow, green, gray, purple, 
                                        red, orange, paleYellow), 
                                      size = 8, replace = FALSE), 
                      guide = guide_legend(nrow = 4, ncol = 2)) +
    theme(axis.text.x = element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank(), legend.position = c(0.7, 0.7), legend.direction = "vertical", legend.title = element_blank(), panel.background = element_blank())
```

# Figure 1, Example 4: Species-pathway correlation

```{r fig1eg4, fig.height=8, fig.width=8}
## update
eh <- ExperimentHub()
myquery <- query(eh, "curatedMetagenomicData")
version <- "20170526."

myquery.stool <- myquery[grepl("stool", myquery$title) & grepl("pathabundance", myquery$title) & grepl(version, myquery$title) & !grepl("VatanenT_2016", myquery$title) & !grepl("AsnicarF_2017", myquery$title), ]

eset.list <- lapply(names(myquery.stool), function(x) myquery.stool[[x]])

names(eset.list) <- gsub("-", "_", gsub(version, "", myquery.stool$title))

for (i in 1:length(eset.list)) {
    colnames(eset.list[[i]]) <- paste(names(eset.list)[[i]], colnames(eset.list[[i]]), sep = ".")
    pData(eset.list[[i]]) <- pData(eset.list[[i]])[, !sapply(pData(eset.list[[i]]), function(x) all(is.na(x)))]
    eset.list[[i]]$subjectID <- as.character(eset.list[[i]]$subjectID)
}

for (i in seq_along(eset.list)) {
    eset.list[[i]] <- eset.list[[i]][!grepl("\\|", rownames(eset.list[[i]])), ]
}

pdat <- joinWithRnames(eset.list, FUN = pData)
pdat$study <- sub("\\..+", "", rownames(pdat))
ab <- joinWithRnames(eset.list, FUN = exprs)
ab[is.na(ab)] <- 0
eset_pathway <- ExpressionSet(assayData = as.matrix(ab), phenoData = AnnotatedDataFrame(pdat))

eset_pathway$prev <- as.numeric(exprs(eset_bugs)[grep("s__Prevotella_copri", rownames(exprs(eset_bugs))), ])

cor_est_p <- function(x1, x2) {
    cor <- cor.test(x1, x2)
    c(r = cor$estimate, p = cor$p.value)
}

cors <- t(sapply(featureNames(eset_pathway), function(i) cor_est_p(exprs(eset_pathway)[i, ], eset_pathway$prev)))

feature <- rownames(cors)

cors <- as.data.frame(cors)
cors$feature <- feature
cors <- na.omit(cors)

par(mar = c(10, 10, 10, 10))
qplot(
    x = eset_pathway$prev,
    y = exprs(eset_pathway)[cors$feature[cors$r.cor == max(cors$r.cor)],],
    xlab = "Prevotella copri abundance",
    ylab = "Pathway abundance",
    main = "Inosine 5 phosphate biosynthesis ",
    colour = I(black),
    shape = I(1)
) +
    theme(plot.title = element_text(hjust = 0.5))
```

# Supplemental Figure 1: Health status classification from species abundance

```{r suppfig1, fig.height=8, fig.width=8, eval=TRUE}
dataset_list <-
    c(  "KarlssonFH_2013 (T2D)",
        "LeChatelierE_2013 (Obesity)",
        "NielsenHB_2014 (IBD)",
        "QinJ_2012 (T2D)",
        "QinN_2014 (Cirrhosis)",
        "ZellerG_2014 (CRC)"
      )
class_list <- c("T2D", "obesity", "IBD", "T2D", "cirrhosis", "CRC")
## update
data_list <- c("EH439", "EH547", "EH457", "EH475", "EH541", "EH535")

## update
eh <- ExperimentHub()

set.seed(0);

for (i in 1:length(dataset_list)) {
    taxabund <- eh[[data_list[i]]]
    
    feat <- t(exprs(taxabund))
    feat <- feat[, grep("(s__|unclassified)", colnames(feat))]
    feat <- feat[, -grep("t__", colnames(feat))]

    meta <- pData(taxabund)
    if (i == 2) {
        meta$study_condition[meta$BMI <= 25] = "lean"
        meta$study_condition[meta$BMI >= 30] = "obesity"
    }
    if (i == 6) {
        meta$study_condition[meta$disease_subtype == "smalladenoma"] = "control"
    }
    meta <- meta["study_condition"]
    all <- cbind(feat, meta)
    if (i == 1) {
        all <- subset(all, study_condition != "IGT")
    }
    if (i == 2) {
        all <- subset(all, study_condition != "control")
    }
    if (i == 4) {
        all <- subset(all, !is.na(study_condition))
    }
    if (i == 6) {
        all <- subset(all, study_condition != "adenoma")
    }
    
    assign(
        paste("rf", i, sep = "_"),
        train(
            study_condition ~ .,
            data = all,
            preProc = c("zv", "scale", "center"),
            method = "rf",
            ntree = 500,
            #tuneGrid = expand.grid(.mtry = c(10, 20, 30, 40, 50, 60, 70, 80, 90, 100, 125, 150, 175, 200, 300, 400, 500)),
            tuneGrid = expand.grid(.mtry = c(50)),
            trControl = trainControl(
                method = "repeatedcv",
                number = 10,
                search = "grid",
                summaryFunction = twoClassSummary,
                classProbs = TRUE,
                savePredictions = TRUE
            )
        )
    )
}

for (i in 1:length(dataset_list)) {
    rf <- get(paste("rf", i, sep = "_"))
    
    if (i == 1) {
        plot.roc(
            rf$pred$obs[rf$pred$mtry == rf$bestTune$mtry],
            rf$pred[, class_list[i]][rf$pred$mtry == rf$bestTune$mtry],
            grid = TRUE,
            ci = TRUE,
            xaxs = "i",
            yaxs = "i",
            col = pallet_reduced[i],
            lty = 1
        )
    } else {
        plot.roc(
            rf$pred$obs[rf$pred$mtry == rf$bestTune$mtry],
            rf$pred[, class_list[i]][rf$pred$mtry == rf$bestTune$mtry],
            grid = TRUE,
            ci = TRUE,
            xaxs = "i",
            yaxs = "i",
            col = pallet_reduced[i],
            lty = 1,
            add = TRUE
        )
    }
}

legend(
    "bottomright",
    box.lwd = 0,
    box.col = "transparent",
    bg = "transparent",
    lwd = 2,
    legend = dataset_list,
    col = pallet_reduced    
)
```

# Supplemental Figure 2: Clustering colored by study and disease states

```{r suppfig2, fig.height=8, fig.width=12}
## update
eh <- ExperimentHub()
myquery <- query(eh, "curatedMetagenomicData")
version <- "20170526."

myquery.stool <- myquery[grepl("stool", myquery$title) & grepl("bugs", myquery$title) & grepl(version, myquery$title) & !grepl("VatanenT_2016", myquery$title) & !grepl("AsnicarF_2017", myquery$title), ]

eset.list <- lapply(names(myquery.stool), function(x) myquery.stool[[x]])

names(eset.list) <- gsub("-", "_", gsub(version, "", myquery.stool$title))

for (i in 1:length(eset.list)) {
    colnames(eset.list[[i]]) <- paste(names(eset.list)[[i]], colnames(eset.list[[i]]), sep = ".")
    pData(eset.list[[i]]) <- pData(eset.list[[i]])[,!sapply(pData(eset.list[[i]]), function(x) all(is.na(x)))]
    eset.list[[i]]$subjectID <- as.character(eset.list[[i]]$subjectID)
}

for (i in seq_along(eset.list)) {
    eset.list[[i]] <- eset.list[[i]][grep("t__", rownames(eset.list[[i]]), invert = TRUE),]
    eset.list[[i]] <- eset.list[[i]][grep("s__|_unclassified", rownames(eset.list[[i]]), perl = TRUE),]
}

pdat <- joinWithRnames(eset.list, FUN = pData)
pdat$study <- sub("\\..+", "", rownames(pdat))
ab <- joinWithRnames(eset.list, FUN = exprs)
ab[is.na(ab)] <- 0
eset <- ExpressionSet(assayData = as.matrix(ab), phenoData = AnnotatedDataFrame(pdat))

## update
pseq <- metaphlanToPhyloseq(tax = exprs(eset), metadat = pData(eset), split = ".")

samp <- data.frame(sample_data(pseq))

dist_bray <- distance(pseq, method = "bray")
dist_js <- distance(pseq, method = "jsd")
dist_rjs <- sqrt(dist_js)

ord_bray <- ordinate(pseq, method = "PCoA", distance = dist_bray)

samp$bray_cluster_2 <- factor(pam(dist_bray, k = 2, cluster.only = T))

Prev <- as.numeric(otu_table(pseq)["s__Prevotella_copri", ])
samp$Prevotella_copri <- Prev

pc1 <- ord_bray$vectors[, 1]
pc2 <- ord_bray$vectors[, 2]

otu_tax <- attr(otu_table(pseq), "dimnames")[[1]]
otu_bacteroides <- otu_table(pseq)[grep("s__Bacteroides", otu_tax), ]
sum_bacteroides <- apply(otu_bacteroides, 2, sum)

df_ordinate <- data.frame(pc1, pc2, bact = sum_bacteroides, prev = Prev, bray2 = as.numeric(samp$bray_cluster_2) + 20)
df_bact <- df_ordinate[df_ordinate$bray2 == 21, ]
df_prev <- df_ordinate[df_ordinate$bray2 == 22, ]

samp$study_condition_simplified[samp$study_condition == "cirrhosis"] <- "cirrhosis"
samp$study_condition_simplified[samp$study_condition %in% c("CRC", "adenoma")] <- "CRC/adenoma"
samp$study_condition_simplified[samp$study_condition == "IBD"] <- "IBD"
samp$study_condition_simplified[samp$study_condition == "STEC"] <- "STEC"
samp$study_condition_simplified[samp$study_condition == "T1D"] <- "T1D"
samp$study_condition_simplified[samp$study_condition %in% c("T2D", "IGT")] <- "T2D/IGT"

df_ord_dataset_disease <-
    data.frame(
        pc1,
        pc2,
        disease = samp$study_condition_simplified,
        study = samp$study,
        prev = df_ordinate$prev
    )

df_ord_dataset_disease %>%
    ggplot(aes(
        x = pc1,
        y = pc2,
        shape = disease,
        color = study,
        size = prev
    )) +
    geom_point() +
    coord_fixed() +
    labs(x = "Axis 1", y = "Axis 2", title = "PCoA by study and disease") +
    scale_shape_manual(values = 20:25) +
    guides(
        shape = guide_legend(ncol = 2, title = "Disease"),
	color = guide_legend(ncol = 2, title = "Study"),
        size = guide_legend(ncol = 2, title = "Prevotella copri abundance")
    ) +
    theme(plot.title = element_text(hjust = 0.5))
```

# Supplemental Figure 3. Clustering scores for enterotypes in gut WGS samples

```{r suppfig3}
pam_bray <- sapply(2:10, function(i) pam(dist_bray, k = i, cluster.only = T))
pam_js <- sapply(2:10, function(i) pam(dist_js, k = i, cluster.only = T))
pam_rjs <- sapply(2:10, function(i) pam(dist_rjs, k = i, cluster.only = T))

plot_cluster_validation = function(bray, js, rjs, legend = T, ...) {
    plot(2:10, bray, type = "b", pch = 1, xlab = "Number of Clusters", ...)
    lines(2:10, js, type = "b", pch = 2, lty = 2)
    lines(2:10, rjs, type = "b", pch = 22, lty = 3)
    if (legend) {
        legend("topright", legend = c("Bray-Curtis", "Jensen-Shannon", "Root Jensen-Shannon"), pch = c(1, 2, 22), lty = 1:3)
    }
}

ch_bray <- apply(pam_bray, 2, function(i) cluster.stats(dist_bray, i)$ch)
ch_js <- apply(pam_js, 2, function(i) cluster.stats(dist_js, i)$ch)
ch_rjs <- apply(pam_rjs, 2, function(i) cluster.stats(dist_rjs, i)$ch)

plot_cluster_validation(ch_bray, ch_js, ch_rjs, legend = T, ylab = "Calinski-Harabasz score", ylim = c(0, 450))

si_bray <- apply(pam_bray, 2, function(i) mean(silhouette(i, dist_bray)[, 3]))
si_js <- apply(pam_js, 2, function(i) mean(silhouette(i, dist_js)[, 3]))
si_rjs <- apply(pam_rjs, 2, function(i) mean(silhouette(i, dist_rjs)[, 3]))

plot_cluster_validation(si_bray, si_js, si_rjs, legend = F, ylab = "Average silhouette width", ylim = c(0, 1))
abline(0.75, 0, lty = 5, col = darkGray)
abline(0.5, 0, lty = 5, col = darkGray)
abline(0.25, 0, lty = 5, col = darkGray)
text("Strong support", x = 9, y = 0.8, col = darkGray)
text("Moderate support", x = 9, y = 0.6, col = darkGray)
text("Weak and could be artificial", x = 8.5, y = 0.4, col = darkGray)
text("No substantial structure", x = 8.5, y = 0.2, col = darkGray)

ps_bray <- prediction.strength(dist_bray, Gmin = 2, Gmax = 10, clustermethod = pamkCBI)
ps_js <- prediction.strength(dist_js, Gmin = 2, Gmax = 10, clustermethod = pamkCBI)
ps_rjs <- prediction.strength(dist_rjs, Gmin = 2, Gmax = 10, clustermethod = pamkCBI)

plot_cluster_validation(ps_bray$mean.pred[2:10], ps_js$mean.pred[2:10], ps_rjs$mean.pred[2:10], ylab = "Prediction Strength", ylim = c(0, 1.1), legend = F)
abline(0.9, 0, lty = 5, col = darkGray)
abline(0.8, 0, lty = 8, col = darkGray)
text("Strong support", x = 9, y = 1, col = darkGray)
text("Moderate support", x = 9, y = 0.85, col = darkGray)
text("Little or no support", x = 9, y = 0.6, col = darkGray)
```

# Supplemental Figure 4: Top correlations between metabolic pathways and genera

```{r suppfig4, fig.height=8, fig.width=12}
## update
eh <- ExperimentHub()
myquery <- query(eh, "curatedMetagenomicData")
version <- "20170526."

myquery.stool <- myquery[grepl("stool", myquery$title) & grepl("pathabundance", myquery$title) & grepl(version, myquery$title) & !grepl("VatanenT_2016", myquery$title) & !grepl("AsnicarF_2017", myquery$title), ]

eset.list <- lapply(names(myquery.stool), function(x) myquery.stool[[x]])

names(eset.list) <- gsub("-", "_", gsub(version, "", myquery.stool$title))

for (i in 1:length(eset.list)) {
    colnames(eset.list[[i]]) <- paste(names(eset.list)[[i]], colnames(eset.list[[i]]), sep = ".")
    pData(eset.list[[i]]) <- pData(eset.list[[i]])[,!sapply(pData(eset.list[[i]]), function(x) all(is.na(x)))]
    eset.list[[i]]$subjectID <- as.character(eset.list[[i]]$subjectID)
}

for (i in seq_along(eset.list)) {
    eset.list[[i]] <- eset.list[[i]][!grepl("\\|", rownames(eset.list[[i]])), ]
}

pdat <- joinWithRnames(eset.list, FUN = pData)
pdat$study <- sub("\\..+", "", rownames(pdat))
ab <- joinWithRnames(eset.list, FUN = exprs)
ab[is.na(ab)] <- 0
eset_pathway <- ExpressionSet(assayData = as.matrix(ab), phenoData = AnnotatedDataFrame(pdat))

myquery.stool <- myquery[grepl("stool", myquery$title) & grepl("bugs", myquery$title) & grepl(version, myquery$title) & !grepl("VatanenT_2016", myquery$title) & !grepl("AsnicarF_2017", myquery$title), ]

eset.list <- lapply(names(myquery.stool), function(x) myquery.stool[[x]])

names(eset.list) <- gsub("-", "_", gsub(version, "", myquery.stool$title))

for (i in 1:length(eset.list)) {
    colnames(eset.list[[i]]) <- paste(names(eset.list)[[i]], colnames(eset.list[[i]]), sep = ".")
    pData(eset.list[[i]]) <- pData(eset.list[[i]])[,!sapply(pData(eset.list[[i]]), function(x) all(is.na(x)))]
    eset.list[[i]]$subjectID <- as.character(eset.list[[i]]$subjectID)
}

for (i in seq_along(eset.list)) {
    eset.list[[i]] <- eset.list[[i]][grep("t__", rownames(eset.list[[i]]), invert = TRUE),]
    eset.list[[i]] <- eset.list[[i]][grep("s__|_unclassified", rownames(eset.list[[i]]), perl = TRUE),]
}

pdat <- joinWithRnames(eset.list, FUN = pData)
pdat$study <- sub("\\..+", "", rownames(pdat))
ab <- joinWithRnames(eset.list, FUN = exprs)
ab[is.na(ab)] <- 0
eset_bugs <- ExpressionSet(assayData = as.matrix(ab), phenoData = AnnotatedDataFrame(pdat))

pseq <- metaphlanToPhyloseq(tax = exprs(eset_bugs), metadat = pData(eset_bugs), split = ".")

glom_genus <- tax_glom(pseq, taxrank = "Genus")
top20_genus_otu_names = names(sort(taxa_sums(glom_genus), TRUE)[1:20])
top20_genus <- tax_table(glom_genus)[top20_genus_otu_names,"Genus"]
subset_genus <- prune_taxa(top20_genus_otu_names, glom_genus)

max_cor_pathway <- function(y, X, margin=1) {
  cors = apply(X, margin, function(x) cor.test(as.numeric(y), as.numeric(x))$estimate)
  rownames(X)[cors==max(cors, na.rm=TRUE)]
}

exprs_pwy <- exprs(eset_pathway)
exprs_pwy <- exprs_pwy[apply(exprs_pwy, 1, function(i) max(i) != 0), ]

max_pathways <- apply(otu_table(subset_genus), 1,  function(y) max_cor_pathway(y, X=exprs_pwy))

cor_matrix <- cor(t(otu_table(subset_genus)), t(exprs(eset_pathway)[max_pathways,]))
rownames(cor_matrix) = top20_genus[rownames(cor_matrix), 1]
colnames(cor_matrix) = gsub(".", " ", gsub("..", " - ", colnames(cor_matrix), fixed=TRUE), fixed=TRUE)

melted_cors <- melt(cor_matrix)
melted_cors %>%
    ggplot(aes(x=Var1, y=Var2, fill=value)) +
    geom_tile() +
    coord_fixed() +
    scale_fill_gradient2(low = blueGreen, high = purple, mid = gray,
     midpoint = 0, space = "Lab",
     name="Pearson\nCorrelation") +
    theme(axis.text.x = element_text(angle = 45, vjust=1, hjust=1, size=6), axis.text.y=element_text(size=6), 
     legend.text = element_text(size=6) ) +
    labs(x = "Genus", y = "Pathway")
```

# Supplemental Figure 5: Alpha diversity from 22 gut studies

```{r suppfig5}
## update
eh <- ExperimentHub()
myquery <- query(eh, "curatedMetagenomicData")
version <- "20170526."

myquery.stool <- myquery[grepl("stool", myquery$title) & grepl("bugs", myquery$title) & grepl(version, myquery$title), ]

eset.list <- lapply(names(myquery.stool), function(x) myquery.stool[[x]])

names(eset.list) <- gsub("-", "_", gsub(version, "", myquery.stool$title))

for (i in 1:length(eset.list)) {
    colnames(eset.list[[i]]) <- paste(names(eset.list)[[i]], colnames(eset.list[[i]]), sep = ".")
    pData(eset.list[[i]]) <- pData(eset.list[[i]])[, !sapply(pData(eset.list[[i]]), function(x) all(is.na(x)))]
    eset.list[[i]]$subjectID <- as.character(eset.list[[i]]$subjectID)
}

for (i in seq_along(eset.list)) {
    eset.list[[i]] <- eset.list[[i]][grep("t__", rownames(eset.list[[i]]), invert = TRUE), ]
    eset.list[[i]] <- eset.list[[i]][grep("s__|_unclassified", rownames(eset.list[[i]]), perl = TRUE), ]
}

pdat <- joinWithRnames(eset.list, FUN = pData)
pdat$study <- sub("\\..+", "", rownames(pdat))
ab <- joinWithRnames(eset.list, FUN = exprs)
ab[is.na(ab)] <- 0
eset <- ExpressionSet(assayData = as.matrix(ab), phenoData = AnnotatedDataFrame(pdat))

## update
pseq <- metaphlanToPhyloseq(tax = exprs(eset), metadat = pData(eset), split = ".")

alpha <- estimate_richness(pseq, measures = "Shannon")
alpha$study <- sample_data(pseq)$study

alpha %<>%
    group_by(study) %>%
    mutate(median = median(Shannon)) %>%
    arrange(desc(median)) %>%
    ungroup %>%
    mutate(study_num = as.numeric(as.factor(alpha$study)))

box_order <- factor(unique(alpha$study[order(alpha$median)]))

alpha$study <- factor(alpha$study, levels = box_order)

alpha %>%
    ggplot(aes(x = study, y = Shannon, fill = study)) +
    stat_boxplot(geom = "errorbar") +
    geom_boxplot() +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
    guides(fill = guide_legend(title="Study")) +
    labs(x = "", y = "Shannon Alpha Diversity") +
    scale_fill_manual(values = c(pallet[1:11], pallet[1:11]))
```

# Session Info
```{r sessioninfo, cache=FALSE}
sessionInfo()
```
