---
title: "curatedMetagenomicData"
author:
- name: Lucas Schiffer, MPH
  affiliation: 
  - Section of Computational Biomedicine, Boston University School of Medicine,
    Boston, MA, U.S.A.
  email: schifferl@bu.edu
package: curatedMetagenomicData
abstract: >
    The curatedMetagenomicData package provides standardized, curated human
    microbiome data for novel analyses. It includes gene families, marker
    abundance, marker presence, pathway abundance, pathway coverage, and
    relative abundance for samples collected from different body sites. The
    bacterial, fungal, and archaeal taxonomic abundances for each sample were
    calculated with MetaPhlAn3 and metabolic functional potential was calculated
    with HUMAnN3. The manually curated sample metadata and standardized
    metagenomic data are available as (Tree)SummarizedExperiment objects.
output: 
  BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{curatedMetagenomicData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

# Installation

Most users should simply install the current version of `r Biocpkg("curatedMetagenomicData")` from Bioconductor, unless they have good reason not to. Changes to the package are always commited to GitHub first, and it might be desirable to have the absolute latest changes by installing `r Githubpkg("waldronlab/curatedMetagenomicData")` from GitHub; although, this is not recommended for most users. Finally, we have made extensive changes in `r Biocpkg("curatedMetagenomicData")` `3.0.0` and some users might want to revert to the previous version of `r Githubpkg("waldronlab/curatedMetagenomicData")` from GitHub for older analyses. The previous version of `r Githubpkg("waldronlab/curatedMetagenomicData")` from GitHub, `1.20.0`, will remain installable for the foreseeable future.

To install the current version of `r Biocpkg("curatedMetagenomicData")` from Bioconductor, use `r CRANpkg("BiocManager")` as follows.

```{r, eval = FALSE}
BiocManager::install("curatedMetagenomicData")
```

To install the current version of `r Githubpkg("waldronlab/curatedMetagenomicData")` from GitHub, use `r CRANpkg("BiocManager")` as follows.

```{r, eval = FALSE}
BiocManager::install("waldronlab/curatedMetagenomicData", dependencies = TRUE, build_vignettes = TRUE)
```

To install the previous version of `r Githubpkg("waldronlab/curatedMetagenomicData")` from GitHub, use `r CRANpkg("BiocManager")` as follows.

```{r, eval = FALSE}
BiocManager::install("waldronlab/curatedMetagenomicData", dependencies = TRUE, build_vignettes = TRUE, ref = "v1.20.0")
```

# R Packages

To demonstrate the functionality of `r Biocpkg("curatedMetagenomicData")`, `r CRANpkg("dplyr")` and `r CRANpkg("DT")` are needed.

```{r, message = FALSE}
library(curatedMetagenomicData)
library(dplyr)
library(DT)
```

# Sample Metadata

The `r Biocpkg("curatedMetagenomicData")` package contains a `data.frame`, `sampleMetadata`, of manually curated sample metadata to help users understand the nature of studies and samples available prior to returning resources. Beyond this, it serves two purposes: 1) to define `study_name` which is used with the `curatedMetagenomicData()` method to query and return resources, and 2) to define `sample_id` which is used with the `returnSamples()` method to return samples across studies.

To demonstrate, the first ten rows and columns (without any `NA` values) of `sampleMetadata` for the `AsnicarF_2017` study are shown in the table below.

```{r, collapse = TRUE}
sampleMetadata %>%
    filter(study_name == "AsnicarF_2017") %>%
    select(where(~ !any(is.na(.x)))) %>%
    slice(1:10) %>%
    select(1:10) %>%
    datatable(options = list(dom = "t"), extensions = "Responsive")
```

# Data Access

To access curated metagenomic data, users will use the `curatedMetagenomicData()` method both to query and return resources. The first argument `pattern` is a regular expression pattern to look for in the titles of resources available in `r Biocpkg("curatedMetagenomicData")`; `""` will return all resources. The title of each resource is a three part string with "." as a delimiter - the fields are `runDate`, `studyName`, and `dataType`. The `runDate` is the date we created the resource and can mostly be ignored by users because if there is more than one date corresponding to a resource, the most recent one is selected automatically. It would be used if a specific `runDate` was needed, but only a single copy of each resource exists currently.

Multiple resources can be queried or returned with a single call to `curatedMetagenomicData()`, only the titles of resources are returned by default.

```{r, collapse = TRUE}
curatedMetagenomicData("AsnicarF_20.+")
```

When the second argument `dryrun` is set to `FALSE`, a `list` of `SummarizedExperiment` and/or `TreeSummarizedExperiment` objects is returned. When a single resource is requested, a single element `list` is returned.

```{r, collapse = TRUE}
curatedMetagenomicData("AsnicarF_2017.relative_abundance", dryrun = FALSE)
```

When the third argument `counts` is set to `TRUE`, relative abundance proportions are multiplied by read depth and rounded to the nearest integer prior to being returned. Also, when multiple resources are requested, the `list` will contain named elements corresponding to each SummarizedExperiment and/or TreeSummarizedExperiment object.

```{r, collapse = TRUE}
curatedMetagenomicData("AsnicarF_20.+.relative_abundance", dryrun = FALSE, counts = TRUE)
```

# Merge Data

To merge the `list` elements returned from the `curatedMetagenomicData` method into a single `SummarizedExperiment` or `TreeSummarizedExperiment` object, users will use the `mergeData()` method provided elements are the same `dataType`.

```{r, collapse = TRUE}
curatedMetagenomicData("LiJ_20.+.marker_abundance", dryrun = FALSE) %>%
    mergeData()
```

The `mergeData()` method works for every `dataType` and will always return the appropriate data structure (a single `SummarizedExperiment` or `TreeSummarizedExperiment` object).

```{r, collapse = TRUE}
curatedMetagenomicData("LiJ_20.+.pathway_abundance", dryrun = FALSE) %>%
    mergeData()
```

This is useful for analysis across entire studies (e.g. meta-analysis); however, when doing analysis across individual samples (e.g. mega-analysis) the `returnSamples()` method is preferable.

```{r, collapse = TRUE}
curatedMetagenomicData("LiJ_20.+.relative_abundance", dryrun = FALSE) %>%
    mergeData()
```

# Return Samples

To return samples across studies, users will use the `returnSamples()` method along with the `sampleMetadata` `data.frame` subset to include only desired samples and metadata. The subset `sampleMetadata` `data.frame` (as the first argument, `sampleMetadata`) will be used to get the desired resources, the `mergeData()` method will be used to merge them, and the subset `sampleMetadata` `data.frame` will be used again to subset the `SummarizedExperiment` or `TreeSummarizedExperiment` object to include only desired samples and metadata.

The `returnSamples()` method requires a second argument `dataType` (`"gene_families"`, `"marker_abundance"`, `"marker_presence"`, `"pathway_abundance"`, `"pathway_coverage"`, or `"relative_abundance"`) be specified. It is often most convenient to subset the `sampleMetadata` `data.frame` using `r CRANpkg("dplyr")` syntax.

```{r, collapse = TRUE}
sampleMetadata %>%
    filter(age >= 18) %>%
    filter(!is.na(alcohol)) %>%
    filter(body_site == "stool") %>%
    select(where(~ !all(is.na(.x)))) %>%
    returnSamples("relative_abundance")
```

Users should also know that the `counts` argument applies to the `returnSamples()` method as well, and can be 
passed as the third argument. Finally, users should also know that any arbitrary columns added to `sampleMetadata` will be present in the `colData` of the `SummarizedExperiment` or `TreeSummarizedExperiment` object that is returned.

# Example Analysis

To demonstrate the utility of `r Biocpkg("curatedMetagenomicData")`, an example analysis is presented below. However, readers should know analysis is generally beyond the scope of `r Biocpkg("curatedMetagenomicData")` and the analysis presented here is for demonstration purposes alone. It is best to consider the output of `r Biocpkg("curatedMetagenomicData")` as the input for analysis more than anything else.

## R Packages

To demonstrate the utility of `r Biocpkg("curatedMetagenomicData")`, `r CRANpkg("stringr")`, `r Biocpkg("mia")`,  `r Biocpkg("scater")`, and `r CRANpkg("vegan")` packages are needed.

```{r, message = FALSE}
library(stringr)
library(mia)
library(scater)
library(vegan)
```

## Prepare Data

In our hypothetical study examine the association of alcohol consumption and stool microbial composition across all annotated samples in `r Biocpkg("curatedMetagenomicData")`. We will examine the alpha diversity (within subject diversity), beta diversity (between subject diversity), and conclude with a few notes on differential abundance analysis.

### Return Samples

First, as above, we use the `returnSamples()` method to return the relevant samples across all studies available in `r Biocpkg("curatedMetagenomicData")`. We want adults over the age of 18, for whom alcohol consumption status is known, and we only want stool samples. The `select(where...` line below simply removes metadata columns which are all `NA` values - they exist in another study but are all `NA` once subsetting has been done. Lastly, the `"relative_abundance"` `dataType` is requested because it contains the relevant information about microbial composition.

```{r, collapse = TRUE, message = FALSE}
alcoholStudy <-
    filter(sampleMetadata, age >= 18) %>%
    filter(!is.na(alcohol)) %>%
    filter(body_site == "stool") %>%
    select(where(~ !all(is.na(.x)))) %>%
    returnSamples("relative_abundance")
```

### Mutate colData

Most of the values in the `sampleMetadata` `data.frame` (which become `colData`) are in lower snake case (e.g. `snake_case`) and don't look nice in plots. Here, the values of the `alcohol` variable are made into title case using `r CRANpkg("stringr")` so they will look nice in plots.

```{r, collapse = TRUE}
colData(alcoholStudy) <-
    colData(alcoholStudy) %>%
    as.data.frame() %>%
    mutate(alcohol = str_replace_all(alcohol, "no", "No")) %>%
    mutate(alcohol = str_replace_all(alcohol, "yes", "Yes")) %>%
    DataFrame()
```

### Agglomerate Ranks

Next, the `splitByRanks` method from `r Biocpkg("mia")` is used to create alternative experiment for each level of the taxonomic tree (e.g. Genus). This allows for diversity and differential abundance analysis at specific taxonomic levels; with this step complete, our data is ready to analyze.

```{r, collapse = TRUE}
altExps(alcoholStudy) <-
    splitByRanks(alcoholStudy)
```

## Alpha Diversity

Alpha diversity is a measure of the within sample diversity of features (relative abundance proportions here) and seeks to quantify the evenness (i.e. are the amounts of different microbes the same) and richness (i.e. are they are large variety of microbial taxa present). The Shannon index (H') is a commonly used measure of alpha diversity, it's estimated here using the `estimateDiversity()` method from the `r Biocpkg("mia")` package.

To quickly plot the results of alpha diversity estimation, the `plotColData()` method from the `r Biocpkg("scater")` package is used along with `r CRANpkg("ggplot2")` syntax.

```{r, collapse = TRUE, fig.cap = "Alpha Diversity - Shannon Index (H')"}
alcoholStudy %>%
    estimateDiversity(abund_values = "relative_abundance", index = "shannon") %>%
    plotColData(x = "alcohol", y = "shannon", colour_by = "alcohol", shape_by = "alcohol") +
    labs(x = "Alcohol", y = "Alpha Diversity (H')") +
    guides(color = guide_none(), shape = guide_none()) +
    theme(legend.position = "none")
```

The figure suggest that those who consume alcohol have higher Shannon alpha diversity than those who do not consume alcohol; however, the difference does not appear to be significant, at least qualitatively.

## Beta Diversity

Beta diversity is a measure of the between sample diversity of features (relative abundance proportions here) and seeks to quantify the magnitude of differences (or similarity) between every given pair of samples. Below it is accessed by Bray–Curtis Principal Coordinates Analysis (PCoA), which is a linear method, and Uniform Manifold Approximation and Projection (UMAP), which is a non-linear method.

### Bray–Curtis PCoA

To calculate pairwise Bray–Curtis distance for every sample in our study we will use the `runMDS()` method from the `r Biocpkg("scater")` package along with the `vegdist()` method from the `r CRANpkg("vegan")` package.

To quickly plot the results of beta diversity analysis, the `plotReducedDim()` method from the `r Biocpkg("scater")` package is used along with `r CRANpkg("ggplot2")` syntax.

```{r, collapse = TRUE, fig.cap = "Beta Diversity - Bray–Curtis PCoA"}
alcoholStudy %>%
    runMDS(FUN = vegdist, method = "bray", exprs_values = "relative_abundance", altexp = "Genus", name = "BrayCurtis") %>%
    plotReducedDim("BrayCurtis",colour_by = "alcohol", shape_by = "alcohol") +
    labs(x = "PCo 1", y = "PCo 2") +
    guides(colour = guide_legend(title = "Alcohol"), shape = guide_legend(title = "Alcohol")) +
    theme(legend.position = c(0.90, 0.85))
```

### UMAP

To calculate the UMAP coordinates of every sample in our study we will use the `runUMAP()` method from the `r Biocpkg("scater")` package package, as it handles the task in a single line.

To quickly plot the results of beta diversity analysis, the `plotReducedDim()` method from the `r Biocpkg("scater")` package is used along with `r CRANpkg("ggplot2")` syntax is used again.

```{r, collapse = TRUE, fig.cap = "Beta Diversity - UMAP (Uniform Manifold Approximation and Projection)"}
alcoholStudy %>%
    runUMAP(exprs_values = "relative_abundance", altexp = "Genus", name = "UMAP") %>%
    plotReducedDim("UMAP",colour_by = "alcohol", shape_by = "alcohol") +
    labs(x = "UMAP 1", y = "UMAP 2") +
    guides(colour = guide_legend(title = "Alcohol"), shape = guide_legend(title = "Alcohol")) +
    theme(legend.position = c(0.90, 0.85))
```

## Differential Abundance

Next, it would be desirable to establish which microbes are differentially abundant between the two groups (those who consume alcohol, and those who do not). The `r Biocpkg("lefser")` and `r Biocpkg("ANCOMBC")` packages feature excellent methods to perform this tasks; however, code is not included here to avoid including many `Suggests` package - `r Biocpkg("curatedMetagenomicData")` had far too many of these in the the previous version and is now very lean. There is a repository of analyses, [curatedMetagenomicAnalyses](https://github.com/waldronlab/curatedMetagenomicAnalyses), on GitHub and a forthcoming paper that will feature extensive demonstrations of analyses - but for now, the suggestions above will have to suffice.

# Type Conversion

Finally, the `r Biocpkg("curatedMetagenomicData")` package previously had methods for conversion to `phyloseq` class objects, and they have been removed in the latest release. It is likely that some users will still want to do analysis using `r Biocpkg("phyloseq")`, and we would like to help them do so - it is just easier if we don't have to maintain the conversion method ourselves. As such, the `r Biocpkg("mia")` package has a method, `makePhyloseqFromTreeSummarizedExperiment`, that will readily do the conversion - users needing this functionality are advised to use it.

```{r, eval = FALSE}
makePhyloseqFromTreeSummarizedExperiment(alcoholStudy, abund_values = "relative_abundance")
```
