---
title: "cMD4: assembling a cross-study cohort"
author: "curatedMetagenomicData team"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{cMD4: assembling a cross-study cohort}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = TRUE,
                      collapse = TRUE,
                      comment = "#>")
```

# Overview

Starting with `curatedMetagenomicData` 4.0 (cMD4), sample-level profiling
data is served from a published DuckDB catalog of hive-partitioned parquet
files produced by the
[`curatedMetagenomicDataETL`](https://github.com/seandavi/curatedMetagenomicDataETL)
pipeline. The backend swap is intentionally transparent: the cMD3
user-facing functions (`curatedMetagenomicData()`, `returnSamples()`,
`mergeData()`) keep their signatures, so existing user code continues
to work unchanged.

Sample metadata originates from
[`curatedMetagenomicDataCuration`](https://github.com/waldronlab/curatedMetagenomicDataCuration)
and is shipped directly with `curatedMetagenomicData` as two
`data.frame` objects so that the runtime package does not depend on
`curatedMetagenomicDataCuration`:

- `harmonized_meta` -- one row per sample, restricted to columns defined
  in the cMD harmonization schema. Use this for portable, schema-aligned
  cohort assembly.
- `all_meta` -- the same rows as `harmonized_meta` plus study-specific
  columns that have not yet been folded into the schema. Use this when
  inspecting studies that are still under harmonization.

This article walks through the canonical cMD4 cross-study workflow:

1. filter the harmonized metadata to the samples of interest across any
   number of studies,
2. call `returnSamples()` to return a `SummarizedExperiment` whose
   `colData` is the filtered metadata and whose assay is pulled from the
   cMD4 DuckDB catalog on demand.

# Package imports

```{r}
library(curatedMetagenomicData)
library(dplyr)
library(SummarizedExperiment)
```

# Step 1 - Inspect the harmonized metadata

`curatedMetagenomicData` ships two metadata tables built from staged
copies of `curatedMetagenomicDataCuration::makeCombinedMetadata()`:

```{r}
## Schema-aligned harmonized release
dim(harmonized_meta)

## Harmonized release plus under-harmonization columns
dim(all_meta)

colnames(harmonized_meta)[1:12]

# Studies currently carried in the harmonized release
harmonized_meta |>
    count(study_name, sort = TRUE) |>
    head()
```

The harmonized table carries curated clinical, demographic, and
sequencing fields, along with ontology term identifiers for values drawn
from controlled vocabularies.

# Step 2 - Define a cohort across studies

Suppose we want adult, feces-derived, healthy samples from Western
cohorts, pooled across every study that contributes such samples:

```{r}
cohort <- harmonized_meta |>
    filter(
        body_site == "feces",
        !is.na(age_years),
        age_years >= 18,
        disease == "Healthy",
        westernized == "Yes"
    ) |>
    distinct(sample_id, .keep_all = TRUE)

nrow(cohort)
cohort |> count(study_name, sort = TRUE) |> head(10)
```

`cohort` is an ordinary `data.frame` - every harmonized metadata column is
preserved and travels with the cohort downstream.

# Step 3 - Return a SummarizedExperiment

`returnSamples()` takes the filtered metadata and a `dataType`. The
function signature is unchanged from cMD3:

```{r}
se <- returnSamples(cohort, dataType = "relative_abundance")
se
```

The resulting `TreeSummarizedExperiment` has

- `assay(se, "relative_abundance")` - a feature x sample numeric matrix
  pivoted from the long-format parquet view in the cMD4 DuckDB catalog,
- `colData(se)` - the filtered `harmonized_meta` rows, aligned to the
  assay columns by `sample_id`,
- column order matching the rows of the filtered `harmonized_meta` that
  were actually found in the catalog (missing samples are dropped with
  a message),
- `rowTree`, `rowData`, and `counts`/`rownames` handling that match
  cMD3 conventions for `relative_abundance`.

# Step 4 - Use the cohort as usual

Because the return value is a standard `(Tree)SummarizedExperiment`, the
entire Bioconductor microbiome stack applies directly. For example:

```{r}
table(colData(se)$country)
head(rowSums(assay(se)))
```

# Available data types

The cMD4 catalog currently exposes the following data types:

| dataType | source |
|---|---|
| `relative_abundance` | MetaPhlAn clade relative abundance |
| `marker_abundance` | MetaPhlAn marker gene abundance |
| `marker_presence` | MetaPhlAn marker gene presence/absence |
| `marker_rel_ab_w_read_stats` | Marker relative abundance with read-level stats |
| `viral_clusters` | MetaPhlAn viral cluster profiles |

The HUMAnN3-derived data types served by cMD3
(`gene_families`, `pathway_abundance`, `pathway_coverage`) are not yet
available in the cMD4 catalog and will raise an error if requested.
They will be reinstated when added to the ETL output.

# Pointing at a different catalog

The cMD4 backend reads its default URL from the
`curatedMetagenomicData.duckdb_url` option. To point the entire package
at a staging catalog or a local copy, set the option in `.Rprofile` or
at the start of a session:

```{r}
options(curatedMetagenomicData.duckdb_url = "/path/to/cmgd.duckdb")
```

All subsequent calls to `curatedMetagenomicData()` and `returnSamples()`
will use the new location.
